Quantifying the number of *rrn* operons across taxonomic ranks

### 2020-11-04 README.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

---
title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ASVs"
author: "Pat Schloss"
date: "9/11/2020"
output:
  github_document:
    html_preview: false
---

@@

## Questions
* Within a genome, how many distinct sequences of the 16S rRNA gene are present relative to the number of copies per genome? How far apart are these sequences from each other? How does this scale from a genome to kingdoms?
* Within a taxa (any level), how many ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?

* Make sure we have taxonomic data for all of our genomes
* Read FASTA files into R (do it on our own)
* inner_join with tsv file
* group_by / summarize to count number of sequences and copies per genome

>>>>

---
title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs"
author: "Pat Schloss"
date: "9/11/2020"
output:
  github_document:
    html_preview: false
---

@@

## Questions
* Within a genome, how many distinct sequences of the 16S rRNA gene are present relative to the number of copies per genome? How far apart are these sequences from each other? How does this scale from a genome to kingdoms?
* Within a taxa (any level), how many ESVs or ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?

### 2020-11-04 exploratory/2020-09-09-genome-sens-spec.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

---
title: "Analyzing the sensitivity and specificity of ASVs for discriminating between genomes"
author: "Pat Schloss"
date: "9/9/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

@@

```{r}
count_tibble <- read_tsv(here("data/processed/rrnDB.count_tibble"),
												 col_types = "cccd")
```

@@

We see that most genomes actually have more than one copy of the *rrn* operon. I wonder whether those different copies are the same sequence / ASV...

### Determine number of ASVs per genome

Considering most genomes have multiple copes of the *rrn* operon, we need to know whether they all have the same ASV. Otherwise we run the risk of splitting a single genome into multiple ASVs.

```{r}
count_tibble %>% 
	group_by(region, genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(region, n_rrn) %>%
	summarize(med_n_asv = median(n_asv),
						mean_n_asv = mean(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75)) %>%
	filter(n_rrn == 7)

count_tibble %>% 
	group_by(region, genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv, color=region)) + geom_smooth(method="lm")
```

Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ASVs per *rrn* operon.

### Determine whether an ASV is unique to genomes they're found in

Instead of looking at the number of ASVs per genome, we want to see the number of genomes per ASV.

```{r}
count_tibble %>%
	group_by(region, asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n)) %>%
	filter(n_genomes == 1)
```

We see that will full length sequences, that 82% of the ASVs were unique to a genome. For the subregions, about 76% of the ASVs were unique to a genome.

### To be determined...
* Can correct for over representation? 
* Consider analysis at species, genus, family, etc. levels
* Consider looking at more broad definition of an ASV

>>>>

---
title: "Analyzing the sensitivity and specificity of ESVs for discriminating between genomes"
author: "Pat Schloss"
date: "9/9/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

@@

```{r}
count_tibble <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
												 col_types = "cccd")
```

@@

We see that most genomes actually have more than one copy of the *rrn* operon. I wonder whether those different copies are the same sequence / ESV...

### Determine number of ESVs per genome

Considering most genomes have multiple copes of the *rrn* operon, we need to know whether they all have the same ESV. Otherwise we run the risk of splitting a single genome into multiple ESVs.

```{r}
count_tibble %>%
	group_by(region, genome) %>%
	summarize(n_esv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(region, n_rrn) %>%
	summarize(med_n_esv = median(n_esv),
						mean_n_esv = mean(n_esv),
						lq_n_esv = quantile(n_esv, prob=0.25),
						uq_n_esv = quantile(n_esv, prob=0.75)) %>%
	filter(n_rrn == 7)

count_tibble %>%
	group_by(region, genome) %>%
	summarize(n_esv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_esv, color=region)) + geom_smooth(method="lm")
```

Surprisingly (or not!) the number of ESVs increases at a rate of about 2 ESVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ESVs per *rrn* operon.

### Determine whether an ESV is unique to genomes they're found in

Instead of looking at the number of ESVs per genome, we want to see the number of genomes per ESV.

```{r}
count_tibble %>%
	group_by(region, esv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n)) %>%
	filter(n_genomes == 1)
```

We see that will full length sequences, that 82% of the ESVs were unique to a genome. For the subregions, about 76% of the ESVs were unique to a genome.

### To be determined...
* Can correct for over representation?
* Consider analysis at species, genus, family, etc. levels
* Consider looking at more broad definition of an ESV

### 2020-11-04 exploratory/2020-09-29-taxa-representation.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

@@

```{r}
n_taxa_per_rank <- metadata_asv %>%
	filter(region == "v19") %>% 
	group_by(rank, taxon) %>%
	summarize(N = n_distinct(genome_id)) %>%
	summarize(n_1 = n_distinct(taxon),
						n_2 = sum(N >= 2),
						n_3 = sum(N >= 3),
						n_4 = sum(N >= 4),
						n_5 = sum(N >= 5),
						n_10 = sum(N >= 10)
						) %>%
	pivot_longer(-rank, names_to = "n_genomes", values_to="n_taxa") %>%
	mutate(n_genomes = factor(n_genomes,
														levels=c("n_1", "n_2", "n_3", "n_4", "n_5", "n_10")))

>>>>

esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
```

@@

```{r}
n_taxa_per_rank <- metadata_esv %>%
	filter(region == "v19") %>%
	group_by(rank, taxon) %>%
	summarize(N = n_distinct(genome_id)) %>%
	summarize(n_1 = n_distinct(taxon),
						n_2 = sum(N >= 2),
						n_3 = sum(N >= 3),
						n_4 = sum(N >= 4),
						n_5 = sum(N >= 5),
						n_10 = sum(N >= 10)
						) %>%
	pivot_longer(-rank, names_to = "n_genomes", values_to="n_taxa") %>%
	mutate(n_genomes = factor(n_genomes,
														levels=c("n_1", "n_2", "n_3", "n_4", "n_5", "n_10")))

### 2020-11-04 exploratory/2020-10-05-rrn-copy-number.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

```{r}
rank_taxon_rrns <- metadata_asv %>%
	filter(region == "v19") %>%
	group_by(kingdom, phylum, class, order, family, genus, species, genome_id) %>%
	summarize(n_rrns = sum(count), .groups="drop") %>% 
	group_by(kingdom, phylum, class, order, family, genus, species) %>%
	summarize(mean_rrns = mean(n_rrns), .groups="drop") %>% 
	pivot_longer(-mean_rrns, names_to="rank", values_to="taxon") %>%
	drop_na(taxon) %>%
	mutate(rank = factor(rank,
											 levels=c("kingdom", "phylum", "class", "order",
											 				 "family", "genus", "species", "strain"))) %>%
	group_by(rank, taxon) %>%
	summarize(mean_rrns = mean(mean_rrns), .groups="drop")

>>>>

```{r}
rank_taxon_rrns <- metadata_esv %>%
	filter(region == "v19") %>%
	group_by(kingdom, phylum, class, order, family, genus, species, genome_id) %>%
	summarize(n_rrns = sum(count), .groups="drop") %>%
	group_by(kingdom, phylum, class, order, family, genus, species) %>%
	summarize(mean_rrns = mean(n_rrns), .groups="drop") %>%
	pivot_longer(-mean_rrns, names_to="rank", values_to="taxon") %>%
	drop_na(taxon) %>%
	mutate(rank = factor(rank,
											 levels=c("kingdom", "phylum", "class", "order",
											 				 "family", "genus", "species", "strain"))) %>%
	group_by(rank, taxon) %>%
	summarize(mean_rrns = mean(mean_rrns), .groups="drop")
